---
title: "Functional_predictions"
author: "Vanessa Arranz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages and useful functions
```{r setup, include=FALSE}
suppressMessages({
        library("tidyverse")
        library("phyloseq")
        library("rstatix")
        library("vegan")
        library("picante")
        library("kableExtra")
        library("reticulate")
        library("ALDEx2")
        library("data.table")
        library("dplyr")
})
```


# Import FAPROTAX

```{r}
# Import FAPROTAX
faprotax_table <- read.delim("FAPROTAX_1.2.10/functional_otu_table.tsv", row.names = 1)
```

# Subset Species
```{r}
ps_SP

# Extract metadata from the phyloseq object
metadata_species_subset <- data.frame(as(sample_data(ps_SP), "data.frame"))
rownames(metadata_species_subset) <- metadata_species_subset$sample.id

# Subset FAPROTAX table based on sample IDs
subset_species <- metadata_species_subset$sample.id
faprotax_Species <- as.data.frame(faprotax_table[, subset_species])
# Check the subsetted pathway abundance table
head(faprotax_Species)

## Columns doesn't sum up to 1 so we need to normalize
colSums(faprotax_Species)
# Normalize each column to sum to 1
faprotax_Species_normalized <- sweep(faprotax_Species, 2, colSums(faprotax_Species), "/")
colSums(faprotax_Species_normalized)

t_Species_normalized <- as.data.frame(t(faprotax_Species_normalized))
t_Species_normalized_metadata<- rbind(
  metadata_species_subset,
  t_Species_normalized
)
t_Species_normalized$sample.id <- rownames(t_Species_normalized)
write.csv(t_Species_normalized, "FAPROTAX_Species_normalized.csv", row.names = FALSE)

# Remove rows that are all 0
faprotax_species_rel <- faprotax_Species_normalized %>%
  filter(rowSums(faprotax_Species_normalized) > 0) %>%  # Remove rows that are 0
  rownames_to_column(var = "faprotax.function")  # Move function names to a column

# Reshape to long format for plotting
faprotax_species_rel_long <- faprotax_species_rel %>%
  pivot_longer(-faprotax.function, names_to = "sample.id", values_to = "count")

# Add METADATA to the long-format data
plot_data <- faprotax_species_rel_long %>%
  left_join(metadata_species_subset, by = "sample.id")

write.csv(metadata_species_subset, "metadata_species_subset.csv", row.names = FALSE)



# Aggregate data to calculate mean abundances and standard deviation per species
# Convert species to factor
plot_data <- plot_data %>%
  mutate(species = as.factor(species)) %>%
  mutate(faprotax.function = as.factor(faprotax.function))

# Function to calculate standard error
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(na.omit(x)))
}
# Calculate mean
species_means <- aggregate(
  count ~ species + faprotax.function,
  data = plot_data,
  FUN = mean
)
# Calculate standard error
species_se <- aggregate(
  count ~ species + faprotax.function,
  data = plot_data,
  FUN = se
)
# Merge mean and standard error into a single data frame
species_means$se_count <- species_se$count

```

## Identify unique functions for each species
```{r}
arbacia_pathways <- plot_data %>%
  filter(species == "ARBACIA" & count > 0.0001) %>%  # Filter meaningful functions
  pull(faprotax.function) %>%
  unique()

paracentrotus_pathways <- plot_data %>%
  filter(species == "PARACENT" & count > 0.0001) %>%
  pull(faprotax.function) %>%
  unique()

# Compare unique and shared pathways
unique_to_arbacia <- setdiff(arbacia_pathways, paracentrotus_pathways)
unique_to_paracentrotus <- setdiff(paracentrotus_pathways, arbacia_pathways)
shared_pathways <- intersect(arbacia_pathways, paracentrotus_pathways)

cat("Unique to Arbacia:", length(unique_to_arbacia), "\n")
cat("Unique to Paracentrotus:", length(unique_to_paracentrotus), "\n")
cat("Shared Pathways:", length(shared_pathways), "\n")
```

## Bar plot with functions for each species 
```{r}
# Filter functions with counts > 0.0001 in at least one species
filtered_species_means <- species_means %>%
  group_by(faprotax.function) %>%
  filter(any(count > 0.0001)) %>%
  ungroup()

## Order for total abundance of functions
# Calculate total abundance for each function
total_abundance <- aggregate(count ~ faprotax.function, data = filtered_species_means, sum)
# Order functions by total abundance (descending)
ordered_functions <- total_abundance[order(-total_abundance$count), "faprotax.function"]
# Update the factor levels for `faprotax.function`
filtered_species_means$faprotax.function <- factor(
  filtered_species_means$faprotax.function,
  levels = ordered_functions
)

# Create the plot 
p <- ggplot(filtered_species_means, aes(x = faprotax.function, y = count, fill = species)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +  # Bars
  geom_errorbar(
    aes(ymin = count - se_count, ymax = count + se_count),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"  # Error bars in black for visibility
  ) +
  scale_fill_manual(values = c("ARBACIA" = "#985D6D", "PARACENT" = "#C9D887")) +
  #scale_y_continuous(
    #limits = c(0.001, 0.5),  # Set Y-axis limits
    #breaks = c(0.0001, 0.001, 0.01, 0.1, 1),  # Define custom breaks
    #labels = scales::label_number()  # Ensure readable labels
  #) +
  theme_minimal(base_size = 12) +  # Clean base theme
  theme(
    panel.background = element_rect(fill = "white", color = NA),  # White background
    panel.grid.major = element_blank(),  # Remove gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    axis.line = element_line(color = "black"),  # Black X and Y axis lines
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),  # Rotate and adjust X-axis text
    axis.text.y = element_text(size = 10),
    legend.position = "top",
    legend.title = element_blank(),  # Simplify legend
    plot.title = element_text(size = 14, face = "bold"),  # Title styling
    plot.subtitle = element_text(size = 12, face = "italic")  # Subtitle styling
  ) +
  labs(
    title = "Relative Abundance of Predicted Functions by Species",
    subtitle = "Y-axis transformed using log10 scale",
    x = "Function",
    y = "Relative Abundance",
    fill = "Species"
  )

# Tranform the y-axis to log10 scale to show the low abundant functions (set ylim more than 0 otherwise can't tranf log10(0))
p2 <- p + coord_trans(y = "log2", ylim = c(0.0001, 1))+ 
      scale_y_continuous(
        breaks = c(0.0001, 0.001, 0.01, 0.1, 1),  # Define custom breaks
        labels = scales::label_number()
        ) 
```

## Statistical test
```{r}
plot_data
write.csv(plot_data, "Species_functions_data.csv")

functions <- unique(plot_data$faprotax.function)

# Create a data frame to store results
kruskal_results <- data.frame()

for (func in functions) {
  subset_data <- filter(plot_data, faprotax.function == func)
  
  # Perform Kruskal-Wallis test
  kruskal_test <- kruskal.test(count ~ species, data = subset_data)
  
  # Store results
  kruskal_results <- rbind(kruskal_results, 
                           data.frame(
                             Function = func,
                             Statistic = as.numeric(kruskal_test$statistic),
                             P_value = kruskal_test$p.value
                           ))
}

# Adjust p-values for multiple testing (Benjamini-Hochberg)
kruskal_results$P_adj <- p.adjust(kruskal_results$P_value, method = "BH")

write.csv(kruskal_results, "Functional_predictions_statisticaltest/Kruskal_Wallis_Results_Species.csv", row.names = FALSE)
```


# Subset Arbacia ST
```{r}
# Subset the faprotax table based on these sample IDs
faprotax_Arb_TS <- as.data.frame(faprotax_table[, subset_Arb_TS])
# Check the subsetted pathway abundance table
head(faprotax_Arb_TS)

colSums(faprotax_Arb_TS)

# transform to relative abundances
faprotax_Arb_TS_abundance_rel <- faprotax_Arb_TS %>%
  filter(rowSums(faprotax_Arb_TS) > 0) %>%  # remove the rows that ara 0
  mutate(across(where(is.numeric), ~ . / sum(.))) %>% # Divide by total sum
  rownames_to_column(var = "faprotax.function")  # Move pathway_class to a column

# Long-format for ggplot
faprotax_Arb_TS_abundance_rel_long <- faprotax_Arb_TS_abundance_rel %>%
  pivot_longer(-faprotax.function, names_to = "sample_name", values_to = "count")

# add metadata
plot_data_Arb_TS <- faprotax_Arb_TS_abundance_rel_long %>%
  left_join(metadata_Arb_TS_subset, by = "sample_name")

custom_palette2 <- colorRampPalette(brewer.pal(12, "Paired"))(38)

# Taxa bar Plot
plot_data_Arb_TS %>%
  ggplot(aes(x = sample_name, y = count)) +
  geom_bar(aes(fill = faprotax.function), stat = "identity", position = "fill") +
  facet_grid(~ body.tissue, scales = "free_x") +
  scale_y_continuous(name = "Relative abundance", labels = scales::percent) +
  scale_fill_manual(values = custom_palette2) +
  theme(axis.text.x = element_text(angle = 90), 
        axis.text.y = element_text(color = "black"))

```

## Identify unique functions for each body type
```{r}
Arb_celo_pathways <- plot_data_Arb_TS %>%
  filter(body.tissue == "CELO" & count > 0.0001) %>%  # Filter meaningful functions
  pull(faprotax.function) %>%
  unique()

Arb_cf_pathways <- plot_data_Arb_TS %>%
  filter(body.tissue == "CF" & count > 0.0001) %>%  # Filter meaningful functions
  pull(faprotax.function) %>%
  unique()

Arb_caca_pathways <- plot_data_Arb_TS %>%
  filter(body.tissue == "CACA" & count > 0.0001) %>%  # Filter meaningful functions
  pull(faprotax.function) %>%
  unique()

# Compare unique and shared pathways
# Calculate intersections
celo_cf_intersect <- intersect(Arb_celo_pathways, Arb_cf_pathways)
celo_caca_intersect <- intersect(Arb_celo_pathways, Arb_caca_pathways)
cf_caca_intersect <- intersect(Arb_cf_pathways, Arb_caca_pathways)
# Calculate unique elements
unique_celo <- setdiff(Arb_celo_pathways, union(Arb_cf_pathways, Arb_caca_pathways))
unique_cf <- setdiff(Arb_cf_pathways, union(Arb_celo_pathways, Arb_caca_pathways))
unique_caca <- setdiff(Arb_caca_pathways, union(Arb_celo_pathways, Arb_cf_pathways))

# Intersection of all three
all_three_intersect <- Reduce(intersect, list(Arb_celo_pathways, Arb_cf_pathways, Arb_caca_pathways))

```

## Bar plot with functions for each body tissue 
```{r}
# Calculate mean
body_tissue_means <- aggregate(
  count ~ body.tissue + faprotax.function,
  data = plot_data_Arb_TS,
  FUN = mean
)
# Calculate standard error
body_tissue_se <- aggregate(
  count ~ body.tissue + faprotax.function,
  data = plot_data_Arb_TS,
  FUN = se
)
# Merge mean and standard error into a single data frame
body_tissue_means$se_count <- body_tissue_se$count


# Filter functions with counts > 0.0001 in at least one species
filtered_body_tissue_means <- body_tissue_means %>%
  group_by(faprotax.function) %>%
  filter(any(count > 0.0001)) %>%
  ungroup()

## Order for total abundance of functions
# Calculate total abundance for each function
total_abundance_Arb_ST <- aggregate(count ~ faprotax.function, data = filtered_body_tissue_means, sum)
# Order functions by total abundance (descending)
ordered_functions_Arb_ST <- total_abundance_Arb_ST[order(-total_abundance_Arb_ST$count), "faprotax.function"]
# Update the factor levels for `faprotax.function`
filtered_body_tissue_means$faprotax.function <- factor(
  filtered_body_tissue_means$faprotax.function,
  levels = ordered_functions_Arb_ST
)

# Create the plot 
Arb_ST_functions_plot <- ggplot(filtered_body_tissue_means, aes(x = faprotax.function, y = count, fill = body.tissue)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +  # Bars
  geom_errorbar(
    aes(ymin = count - se_count, ymax = count + se_count),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"  # Error bars in black for visibility
  ) +
  scale_fill_manual(values = c(CELO="#F2AD00", CF="#F98400", CACA="#00A08A")) +
  theme_minimal(base_size = 12) +  # Clean base theme
  theme(
    panel.background = element_rect(fill = "white", color = NA),  # White background
    panel.grid.major = element_blank(),  # Remove gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    axis.line = element_line(color = "black"),  # Black X and Y axis lines
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),  # Rotate and adjust X-axis text
    axis.text.y = element_text(size = 10),
    legend.position = "top",
    legend.title = element_blank(),  # Simplify legend
    plot.title = element_text(size = 14, face = "bold"),  # Title styling
    plot.subtitle = element_text(size = 12, face = "italic")  # Subtitle styling
  ) +
  labs(
    title = "Relative Abundance of Predicted Functions by Body types Arbacia",
    subtitle = "Y-axis transformed using log10 scale",
    x = "Function",
    y = "Relative Abundance",
    fill = "Species"
  )

# Tranform the y-axis to log10 scale to show the low abundant functions (set ylim more than 0 otherwise can't tranf log10(0))
Arb_ST_functions_plot <- Arb_ST_functions_plot + coord_trans(y = "log2", ylim = c(0.0001, 1))+ 
      scale_y_continuous(
        breaks = c(0.0001, 0.001, 0.01, 0.1, 1),  # Define custom breaks
        labels = scales::label_number()
        ) 
```

## Statistical test
```{r}
plot_data_Arb_TS
write.csv(plot_data_Arb_TS, "plot_data_Arb_TS.csv")

functions <- unique(plot_data_Arb_TS$faprotax.function)

# Create a data frame to store results
kruskal_results <- data.frame()

for (func in functions) {
  subset_data <- filter(plot_data_Arb_TS, faprotax.function == func)
  
  # Perform Kruskal-Wallis test
  kruskal_test <- kruskal.test(count ~ body.tissue, data = subset_data)
  
  # Store results
  kruskal_results <- rbind(kruskal_results, 
                           data.frame(
                             Function = func,
                             Statistic = as.numeric(kruskal_test$statistic),
                             P_value = kruskal_test$p.value
                           ))
}

# Adjust p-values for multiple testing (Benjamini-Hochberg)
kruskal_results$P_adj <- p.adjust(kruskal_results$P_value, method = "bonferroni")

wilcoxon_results <- data.frame()

for (func in functions) {
  # Check if the Kruskal-Wallis test is significant
  if (kruskal_results$P_value[kruskal_results$Function == func] < 0.05) {
    subset_data <- filter(plot_data_Arb_TS, faprotax.function == func)
    
    # Perform pairwise Wilcoxon test
    pairwise_test <- pairwise.wilcox.test(subset_data$count, 
                                          subset_data$body.tissue, 
                                          p.adjust.method = "BH")
    
    # Convert pairwise results into a data frame
    pairwise_df <- as.data.frame(as.table(pairwise_test$p.value))
    colnames(pairwise_df) <- c("Group1", "Group2", "P_value")
    pairwise_df$Function <- func
    
    # Append to results
    wilcoxon_results <- rbind(wilcoxon_results, pairwise_df)
  }
}

write.csv(kruskal_results, "Functional_predictions_statisticaltest/Kruskal_Wallis_Results_Arb_ST.csv", row.names = FALSE)
write.csv(wilcoxon_results, "Functional_predictions_statisticaltest/Pairwise_Wilcoxon_Results_Arb_ST.csv", row.names = FALSE)
```

# Subset Arbacia Location

```{r}
# Extract metadata from the phyloseq object
metadata_Arb_LOC_subset <- data.frame(as(sample_data(Arb_LOC), "data.frame"))
rownames(metadata_Arb_LOC_subset) <- metadata_Arb_LOC_subset$sample.id

# Subset FAPROTAX table based on sample IDs
subset_Arb_LOC <- metadata_Arb_LOC_subset$sample.id
faprotax_Arb_LOC <- as.data.frame(faprotax_table[, subset_Arb_LOC])
# Check the subsetted pathway abundance table
head(faprotax_Arb_LOC)

## Columns doesn't sum up to 1 so we need to normalize
colSums(faprotax_Arb_LOC)
# Normalize each column to sum to 1
faprotax_Arb_LOC <- sweep(faprotax_Arb_LOC, 2, colSums(faprotax_Arb_LOC), "/")
colSums(faprotax_Arb_LOC)

# Remove rows that are all 0
faprotax_Arb_LOC_rel <- faprotax_Arb_LOC %>%
  filter(rowSums(faprotax_Arb_LOC) > 0) %>%  # Remove rows that are 0
  rownames_to_column(var = "faprotax.function")  # Move function names to a column

# Reshape to long format for plotting
faprotax_Arb_LOC_rel_long <- faprotax_Arb_LOC_rel %>%
  pivot_longer(-faprotax.function, names_to = "sample.id", values_to = "count")

# Add METADATA to the long-format data
plot_data_Arb_LOC <- faprotax_Arb_LOC_rel_long %>%
  left_join(metadata_Arb_LOC_subset, by = "sample.id")

# Aggregate data to calculate mean abundances and standard deviation per species
# Convert species to factor
plot_data_Arb_LOC <- plot_data_Arb_LOC %>%
  mutate(place = as.factor(place)) %>%
  mutate(faprotax.function = as.factor(faprotax.function))

# Function to calculate standard error
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(na.omit(x)))
}
# Calculate mean
Arb_LOC_means <- aggregate(
  count ~ place + faprotax.function,
  data = plot_data_Arb_LOC,
  FUN = mean
)
# Calculate standard error
Arb_LOC_se <- aggregate(
  count ~ place + faprotax.function,
  data = plot_data_Arb_LOC,
  FUN = se
)
# Merge mean and standard error into a single data frame
Arb_LOC_means$se_count <- Arb_LOC_se$count

# Stacked bar plot: Mean functional abundances by species
custom_palette2 <- colorRampPalette(brewer.pal(12, "Paired"))(38)

barplot_functions_Arb_LOC <- ggplot(Arb_LOC_means, aes(x = place, y = count, fill = faprotax.function)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(name = "Relative Abundance", labels = scales::percent) +
  scale_fill_manual(values = custom_palette2) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right") +
  labs(title = "FAPROTAX Functional Predictions in Arbacia by Location", x = "Location")

```

## Identify unique functions for each species
```{r}
Arb_blanes_pathways <- plot_data_Arb_LOC %>%
  filter(place == "BLANES" & count > 0.0001) %>%  # Filter meaningful functions
  pull(faprotax.function) %>%
  unique()

Arb_palma_pathways <- plot_data_Arb_LOC %>%
  filter(place == "CONTROL" & count > 0.0001) %>%  # Filter meaningful functions
  pull(faprotax.function) %>%
  unique()


# Compare unique and shared pathways
unique_to_arb_blanes <- setdiff(Arb_blanes_pathways, Arb_palma_pathways)
unique_to_arb_palma <- setdiff(Arb_palma_pathways, Arb_blanes_pathways)
shared_pathways <- intersect(Arb_blanes_pathways, Arb_blanes_pathways)

cat("Unique to Arbacia Blanes:", length(unique_to_arb_blanes), "\n")
cat("Unique to Arbacia Palma:", length(unique_to_arb_palma), "\n")
cat("Shared Pathways:", length(shared_pathways), "\n")
```

## Bar plot with functions for each location
```{r}
# Filter functions with counts > 0.0001 in at least one location
filtered_Arb_LOC_means <- Arb_LOC_means %>%
  group_by(faprotax.function) %>%
  filter(any(count > 0.0001)) %>%
  ungroup()

## Order for total abundance of functions
# Calculate total abundance for each function
total_abundance_Arb_LOC <- aggregate(count ~ faprotax.function, data = filtered_Arb_LOC_means, sum)
# Order functions by total abundance (descending)
ordered_functions_Arb_LOC <- total_abundance_Arb_LOC[order(-total_abundance_Arb_LOC$count), "faprotax.function"]
# Update the factor levels for `faprotax.function`
filtered_Arb_LOC_means$faprotax.function <- factor(
  filtered_Arb_LOC_means$faprotax.function,
  levels = ordered_functions_Arb_LOC)

# Create the plot 
Arb_LOC_functions_plot <- ggplot(filtered_Arb_LOC_means, aes(x = faprotax.function, y = count, fill = place)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +  # Bars
  geom_errorbar(
    aes(ymin = count - se_count, ymax = count + se_count),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black"  # Error bars in black for visibility
  ) +
  scale_fill_manual(values = c("BLANES" = "#FD6467", "CONTROL" = "#5B1A18")) +
  theme_minimal(base_size = 12) +  # Clean base theme
  theme(
    panel.background = element_rect(fill = "white", color = NA),  # White background
    panel.grid.major = element_blank(),  # Remove gridlines
    panel.grid.minor = element_blank(),  # Remove minor gridlines
    axis.line = element_line(color = "black"),  # Black X and Y axis lines
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),  # Rotate and adjust X-axis text
    axis.text.y = element_text(size = 10),
    legend.position = "top",
    legend.title = element_blank(),  # Simplify legend
    plot.title = element_text(size = 14, face = "bold"),  # Title styling
    plot.subtitle = element_text(size = 12, face = "italic")  # Subtitle styling
  ) +
  labs(
    title = "Relative Abundance of Predicted Functions by Location in Arbacia",
    subtitle = "Y-axis transformed using log10 scale",
    x = "Function",
    y = "Relative Abundance",
    fill = "Species"
  )

# Tranform the y-axis to log10 scale to show the low abundant functions (set ylim more than 0 otherwise can't tranf log10(0))
Arb_LOC_functions_plot <- Arb_LOC_functions_plot + coord_trans(y = "log2", ylim = c(0.0001, 1))+ 
      scale_y_continuous(
        breaks = c(0.0001, 0.001, 0.01, 0.1, 1),  # Define custom breaks
        labels = scales::label_number()
        ) 
```
